Methods that remove batch effects while retaining group differences may lead to exaggerated confidence in downstream analyses

### 2014-07-22 tex/manuscript.tex
div

Our knowledge of the problem discussed in this article came through a typical use case when trying to adjust for batch effects in an unbalanced data set using ComBat. Upon realizing that the confidence in the estimated group differences was exaggerated, we searched the literature for a better understanding of correct use and the potential limitations of ComBat. However, the authors of ComBat and the \texttt{sva} package recommended including study group as a covariate as we had done. In addition, other studies investigating batch effects were mostly recommending ComBat without much concern (\citealp{Kupfer2012}, \citealp{Kitchen2011}, \CITE{PMID:21386892}). A brief inquiry into  some of the articles citing ComBat (574 on Google Scholar) revealed few reported problems, although their method descriptions regarding ComBat were mostly sparse, limited to one or two sentences. A further indication of the carefree use of the procedure was the frequent omittance of program parameters that were applied, i.e batch labels or whether group labels were supplied as covariates. Often, no effort was undertaken in order to substantiate the existence of batch effects, beyond stating the presence of batches. The incorporation of ComBat into analysis tools, like \COMMENT{TCGA}{does not seem to offer ComBat with covariates} and GenePattern (\citealp{Reich2006}), and other packages like \texttt{sva} (\citealp{Leek2012}) and \COMMENT{inSilicoDb (\citealp{Taminau2011})}{does not seem to offer ComBat with covariates} could make it more accessible, but its use less transparent. We are concerned, in light of our findings, that many published results from data adjusted by ComBat with study group as covariate may be unreliable. Furthermore, the frequent lack of proper method description accompanying published results makes it hard to judge if they are affected or not. We hope that our warning will help caution the scientific community against this particular approach.

>>>>

Our knowledge of the problem discussed in this article came through a typical use case when trying to adjust for batch effects in an unbalanced data set using ComBat. Upon realizing that the confidence in the estimated group differences was exaggerated, we searched the literature for a better understanding of correct use and the potential limitations of ComBat. However, the authors of ComBat and the \texttt{sva} package recommended including study group as a covariate as we had done. In addition, other studies investigating batch effects were mostly recommending ComBat without much concern (\citealp{Kupfer2012}, \citealp{Kitchen2011}, \CITE{PMID:21386892}). A brief inquiry into  some of the articles citing ComBat (574 on Google Scholar) revealed few reported problems, although their method descriptions regarding ComBat were mostly sparse, limited to one or two sentences. A further indication of the carefree use of the procedure was the frequent omittance of program parameters that were applied, i.e batch labels or whether group labels were supplied as covariates. Often, no effort was undertaken in order to substantiate the existence of batch effects, beyond stating the presence of batches. The incorporation of ComBat into analysis tools, like \COMMENT{TCGA}{does not seem to offer ComBat with covariates} and GenePattern (\citealp{Reich2006}), and other packages like \COMMENT{inSilicoDb (\citealp{Taminau2011})}{does not seem to offer ComBat with covariates} could make it more accessible, but its use less transparent. We are concerned, in light of our findings, that many published results from data adjusted by ComBat with study group as covariate may be unreliable. Furthermore, the frequent lack of proper method description accompanying published results makes it hard to judge if they are affected or not. We hope that our warning will help caution the scientific community against this particular approach.
